\pkg{DNest4}: Diffusive Nested Sampling in \proglang{C++} and \proglang{Python}

### 2016-05-26 paper/ms.tex
log likelihood and print
Caticha and Giffin

The log-likelihood code is:
\begin{CodeChunk}
\begin{CodeInput}
double StraightLine::log_likelihood() const
{
    // Grab the dataset
    const std::vector<double>& x = Data::get_instance().get_x();
    const std::vector<double>& y = Data::get_instance().get_y();

@@

std::string StraightLine::description() const
{
    return std::string("m, b, sigma");
}

\end{CodeInput}
\end{CodeChunk}

@@

A simple ``rejection sampling'' version of
ABC works by sampling the joint prior distribution for the parameters and
data
\begin{align}
p(\theta, D | I) &= p(\theta | I)p(D | \theta, I)
\end{align}
and rejecting samples for which $D \neq D_{\rm observed}$, so that the
samples represent
\begin{align}
p(\theta, D | D = D_{\rm observed}, I) &\propto p(\theta | I)p(D | \theta, I)
\mathds{1}\left[D = D_{\rm observed}\right].
\end{align}
The marginal distribution for $\theta$ is also the conditional distribution
$p(\theta | D = D_{\rm observed}, I)$, that is, the posterior.

@@

\begin{table}[ht!]
\centering
\small
\begin{tabular}{|l|l|l|}
\hline
Bayesian computation		&		Statistical mechanics		&		ABC\\
\hline
Parameter space	$\Theta$	&		Phase space	$\Omega$ 			& Product space $\Theta \times \mathcal{D}$\\
Posterior $p(\params | D,I)$   &  Canonical distribution $p(\boldsymbol{x}; T)$     & \\ 
Marginal likelihood $P(D|I)$	&	Partition function $Z(T)$	&   Marginal likelihood $p(D\in | I)$\\
\hline
\end{tabular}
\caption{\it The relationship between standard Bayesian computation, statistical
mechanics, and ABC. In each case Monte Carlo methods are used to calculate
integrals over a space. In standard Bayesian computation, it is the parameter
space of a model that is usually of interest, whereas in ABC it is the space
of possible parameter values {\it and} data sets.
\label{tab:relation}}
\end{table}

\begin{figure}[ht!]
\centering
\includegraphics[scale=0.5]{figures/joint.pdf}
\caption{\it The joint prior for parameters $\params$ and data $\data$.
\label{fig:joint}}
\end{figure}

The main challenges associated with ABC are:
\begin{enumerate}[(a)]
\item The choice of summary statistics
\item The choice of distance function
\item How to choose the value of $\epsilon$, or how to change it as a run
progresses
\item How to achieve good mixing.
\end{enumerate}
Challenges 1 and 2 are Bayesian in nature, i.e. they relate to the very
definition of the posterior distribution and are well defined independent of
the fact that we are going to use a Monte Carlo method to compute the
results. On the other hand, challenges 3 and 4 are about the computational implementation.

\section{Parameterising the data space}
In standard MCMC-based ABC, proposal moves are made by first proposing a
change to the parameters (from $\params$ to $\params'$), and then generating
a mock dataset $\data'$ from $p(\data | \params')$. As a Metropolis proposal
for exploring the product space $\Theta \times \mathcal{D}$ this is a very
aggressive choice, since the proposed value of the
dataset $\params'$ is likely to be very different from the previous one.
One of the most elementary pieces of advice we receive when learning the
Metropolis algorithm is to be careful about how ambitious our proposal
distributions are, since tiny and huge proposal moves are both very
inefficient. The performance of MCMC-based ABC can therefore be improved by
alternative choices of proposal distribution.

In some cases, we might be able to find efficient proposal distributions for exploring $\Theta \times \mathcal{D}$ by having insight into the particular
problem we're working on. However, there are also techniques that are
completely general in that they should work on any ABC-type problem.

Consider the process of generating a mock dataset. Whatever else is involved,
this process will involve calls to an underlying random number generator.
Perhaps a function {\tt rand()}, generating variables from a Uniform(0, 1)
distribution, is called $n$ times in the course of generating a dataset.
Usually, the simulation process has the property that, if $\theta'$ is
close to $\theta$, then the same sequence of random numbers returned by
{\tt rand()} would produce a dataset $\data'$ that is very similar to $\data$.
This suggests an alternative parameterisation of
the product space $\Theta\times\mathcal{D}$: instead of using the one that
naturally describes the data, we can use a coordinate system defined by
$\theta$ together with $\{u_1, ..., u_n\}$, the $n$ uniform random numbers
used to construct a mock dataset.

In this alternative coordinate system, a proposed move
to change $\params$ while keeping
$\{u_i\}$ fixed results in a proposed new dataset that is similar to the
previous one, assuming the change to $\params$ is small. The standard procedure
of generating an entirely new mock dataset each step is now viewed as a
proposal where $\params$ is moved {\it and all of the $u_i$ coordinates are
re-generated from their iid uniform prior}.

@@

@@

\bibitem[Caticha(2008)]{caticha} Caticha, A.\ 2008.\ Lectures 
on Probability, Entropy, and Statistical Physics.\ ArXiv e-prints 
arXiv:0808.0012.

@@

>>>>

The log-likelihood code is standard:
\begin{CodeChunk}
\begin{CodeInput}
double StraightLine::log_likelihood() const
{
    // Grab the dataset
    const std::vector<double>& x = Data::get_instance().get_x();
    const std::vector<double>& y = Data::get_instance().get_y();

@@

std::string StraightLine::description() const
{
    return std::string("m, b, sigma");
}
\end{CodeInput}
\end{CodeChunk}
The print function specifies that the parameters
$m$, $b$, and $\sigma$ are printed a single line
(of {\tt sample.txt}) and are separated by spaces (the
\code{postprocess} function assumes this is the case).

@@

A simple ``rejection sampling'' version of
ABC works by sampling the joint prior distribution for the parameters and
data
\begin{align}
p(\theta, D | I) &= p(\theta | I)p(D | \theta, I)
\end{align}
and rejecting samples for which $D \neq D_{\rm observed}$, so that the
samples represent
\begin{align}
p(\theta, D | D = D_{\rm observed}, I) &\propto p(\theta | I)p(D | \theta, I)
\mathds{1}\left[D = D_{\rm observed}\right].
\end{align}
The marginal distribution for $\theta$ is also the conditional distribution
$p(\theta | D = D_{\rm observed}, I)$, that is,
the posterior\footnote{Incidentally, this usage of the {\em joint} posterior
distribution for the parameters {\em and the data} provides a bridge
connecting Bayesian updating to maximum entropy updating \citep{caticha,giffin}.}.

@@

%\begin{table}[ht!]
%\centering
%\small
%\begin{tabular}{|l|l|l|}
%\hline
%Bayesian computation		&		Statistical mechanics		&		ABC\\
%\hline
%Parameter space	$\Theta$	&		Phase space	$\Omega$ 			& Product space $\Theta \times \mathcal{D}$\\
%Posterior $p(\params | D,I)$   &  Canonical distribution $p(\boldsymbol{x}; T)$     & \\ 
%Marginal likelihood $P(D|I)$	&	Partition function $Z(T)$	&   Marginal likelihood $p(D\in | I)$\\
%\hline
%\end{tabular}
%\caption{\it The relationship between standard Bayesian computation, statistical
%mechanics, and ABC. In each case Monte Carlo methods are used to calculate
%integrals over a space. In standard Bayesian computation, it is the parameter
%space of a model that is usually of interest, whereas in ABC it is the space
%of possible parameter values {\it and} data sets.
%\label{tab:relation}}
%\end{table}

%\begin{figure}[ht!]
%\centering
%\includegraphics[scale=0.5]{figures/joint.pdf}
%\caption{\it The joint prior for parameters $\params$ and data $\data$.
%\label{fig:joint}}
%\end{figure}

%The main challenges associated with ABC are:
%\begin{enumerate}[(a)]
%\item The choice of summary statistics
%\item The choice of distance function
%\item How to choose the value of $\epsilon$, or how to change it as a run
%progresses
%\item How to achieve good mixing.
%\end{enumerate}
%Challenges 1 and 2 are Bayesian in nature, i.e. they relate to the very
%definition of the posterior distribution and are well defined independent of
%the fact that we are going to use a Monte Carlo method to compute the
%results. On the other hand, challenges 3 and 4 are about the computational implementation.

%\section{Parameterising the data space}
%In standard MCMC-based ABC, proposal moves are made by first proposing a
%change to the parameters (from $\params$ to $\params'$), and then generating
%a mock dataset $\data'$ from $p(\data | \params')$. As a Metropolis proposal
%for exploring the product space $\Theta \times \mathcal{D}$ this is a very
%aggressive choice, since the proposed value of the
%dataset $\params'$ is likely to be very different from the previous one.
%One of the most elementary pieces of advice we receive when learning the
%Metropolis algorithm is to be careful about how ambitious our proposal
%distributions are, since tiny and huge proposal moves are both very
%inefficient. The performance of MCMC-based ABC can therefore be improved by
%alternative choices of proposal distribution.

%In some cases, we might be able to find efficient proposal distributions for exploring $\Theta \times \mathcal{D}$ by having insight into the particular
%problem we're working on. However, there are also techniques that are
%completely general in that they should work on any ABC-type problem.

%Consider the process of generating a mock dataset. Whatever else is involved,
%this process will involve calls to an underlying random number generator.
%Perhaps a function {\tt rand()}, generating variables from a Uniform(0, 1)
%distribution, is called $n$ times in the course of generating a dataset.
%Usually, the simulation process has the property that, if $\theta'$ is
%close to $\theta$, then the same sequence of random numbers returned by
%{\tt rand()} would produce a dataset $\data'$ that is very similar to $\data$.
%This suggests an alternative parameterisation of
%the product space $\Theta\times\mathcal{D}$: instead of using the one that
%naturally describes the data, we can use a coordinate system defined by
%$\theta$ together with $\{u_1, ..., u_n\}$, the $n$ uniform random numbers
%used to construct a mock dataset.

%In this alternative coordinate system, a proposed move
%to change $\params$ while keeping
%$\{u_i\}$ fixed results in a proposed new dataset that is similar to the
%previous one, assuming the change to $\params$ is small. The standard procedure
%of generating an entirely new mock dataset each step is now viewed as a
%proposal where $\params$ is moved {\it and all of the $u_i$ coordinates are
%re-generated from their iid uniform prior}.

@@

{\bf To be written by Dan FM}

@@

\bibitem[Caticha and Giffin(2006)]{caticha} Caticha, A., Giffin, A.\ 2006.\ Updating Probabilities.\ Bayesian Inference and Maximum Entropy Methods In Science and Engineering 872, 31-42.

@@

\bibitem[Giffin and Caticha(2007)]{giffin} Giffin, A., Caticha, A.\ 2007.\ Updating Probabilities with Data and Moments.\ Bayesian Inference and Maximum Entropy Methods in Science and Engineering 954, 74-84.
